---
title: "Week5 - Spatial Points"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reprojecting Raster data and plotting it with ggplot2

### Quick tutorial - ggplot2 themes
Ever notice the gray background on ggplot2? Well that's referred to as the default theme. There are other themes that can be called every time you createa a ggplot. Let's take a look:

The black & white theme
```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors", x="Year", caption="National Park Database") +
    theme_bw()
  
```

The dark theme

```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors", x="Year", caption="National Park Database") +
    theme_dark()
  
```

The light theme

```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors", x="Year", caption="National Park Database") +
    theme_light()
  
```

The classic theme

```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors",x="Year", caption="National Park Database") +
    theme_classic()
  
```

The void theme

```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors",x="Year", caption="National Park Database") +
    theme_void()
  
```



# Advanced Plotting - Synthesizing Shapefiles 

Often times, the spatial data that we want aren't in a nice, gridded lat/lon. The oblate spheroid that we call home presents some challenges when it comes to displaying spatial datasets. Often times, datasets like these are stored as separate entities - shapefiles & data tables. Fortunately, we have multiple avenues for working with this data. Let's check out a coastal flooding product that projects potential water levels above normal for each day. [coast-flood.udel.edu]




```{r}
#Load packages
library(RColorBrewer)
library(rgdal)
library(sp)
library(ggplot2)
library(ggmap)
library(scales)
library(viridis)

# Use the readOGR function to open a shapefile & constituents
coast.shp <- readOGR("/Users/james/Documents/Github/geog473-673/datasets/cfms_shapefiles/cfms_watersheds.shp")
class(coast.shp)
coast.shp@proj4string
```
This projection is **unique**. It's a transverse mercator with some specific lat_0 and lon_0 starting points. Notice the class of the shapefile - the underlying package controlling it's translation to R is the `sp` package -  A package providing classes and methods for spatial data: points, lines, polygons and grids. 

NOTE: Everytime you read in a shapefile like `cfms_watersheds.shp`, you MUST have the `.shx`, `.prj`, `.dbf`, etc. all within the SAME folder. Even though we are only keying an "opening" of the `.shp` file, `readOGR` is opening them all. 

ANOTHER NOTE: Notice the *@* symbol in `coast.shp@proj4string`. This is you query metadata associated with shapefiles. 

Back to plotting...



```{r}
plot(coast.shp)
# Open the dataset that corresponds to the water levels within the shapefile boxes 
coast.data <- read.csv("/Users/james/Documents/Github/geog473-673/datasets/cfms_shapefiles/water_levels.csv")
head(coast.data)
# now let's find the matching key - in this case, the matching key is the "station".
coast.shp$station
coast.data$station

# notice the difference above - let's reorder the data from the shapefile and the csv data by the station, otherwise
# merging will NOT work. 
coast.shp  = coast.shp[order(as.vector(coast.shp$station)),]
coast.data  = coast.data[order(as.vector(coast.data$station)),]

# merge together the shapefile data and the csv data based on common variable, here called 'station' - we MUST use the duplicateGeoms argument because there are multiple values for the same station name
merged.coast = sp::merge(coast.shp,coast.data,by='station', duplicateGeoms = TRUE)
class(merged.coast)
# remember that package::function is another way to call a specific function. I did this becuase other packages have 'merge' functions
# and I didn't want them to be confused. 

# Now let's use brewer.pal from RColorBrewer to create a blues color pallette
mycolours <- brewer.pal(8, "Blues")
mybreaks <- seq(0,6,0.5) 
# the data we're concerned with here is shortnamed maxpred - maximum predicted water level for the next 24 hours
cut(merged.coast$maxpred, mybreaks)
mycolourscheme <- mycolours[findInterval(merged.coast$maxpred, vec = mybreaks)]

```



The data is prepared, now we just need to plug in everything to plotting functions. First up is `spplot` from the `sp` package:



```{r}
spplot(merged.coast, "maxpred", par.settings = list(axis.line = list(col ="transparent")), main = "Projected Water Levels (Feet)", cuts = 5, col ="transparent", col.regions = mycolours)

```



Now we use the `tmap` function - very similar to `ggplot2`. `tmap` == Thematic Map Visualization. Thematic maps are geographical maps in which spatial data distributions are visualized. This package offers a flexible, layer-based, and easy to use approach to create thematic maps, such as choropleths and bubble maps. It is based on the grammar of graphics, and resembles the syntax of ggplot2.




```{r}
library(tmap)
tm_shape(merged.coast) + 
  tm_polygons(col='maxpred', title = "Projected Water Levels", palette = "Spectral") + tm_style("classic") + tm_scale_bar(position = c("left", "bottom")) 
```



Ok, so those are cool, but what about using `ggplot2`? In order to do this, we'll need to use `geom_sf` from `ggplot2`. This enables us to plot `sf` objects. `sf` is a package for Simple Features. It's one of the other big players for spatial data wrangling in R that a lot of other, fancier packages that we use in this class are built on. We'll need to load in `sf` before we can convert our shapefile into an `sf` object. 



```{r}
library(sf)
states <- readOGR("/Users/james/Documents/Github/geog473-673/datasets/ne_10m_admin_1_states_provinces/ne_10m_admin_1_states_provinces.shp")
class(states)
states = st_as_sf(states)
class(states)

# convert it to an sf object
gg_merged = st_as_sf(merged.coast)
class(gg_merged)
# now we use geom_sf since our merged.coast is now a simple feature
ggplot(data = states) + geom_sf() + theme_void() +
  geom_sf(data = gg_merged, aes(fill = maxpred)) +
  coord_sf(xlim = c(-76.1, -74.7), ylim = c(38.3,40), expand = FALSE) +
  scale_fill_distiller(palette = "Blues", direction= 1)

```


Looks pretty good even though the shapefile isn't quite as sharp as we'd like. Besides the shapefile which we added to the plot via `geom_sf()`, take notice of the `theme_void()`...this plot will NOT WORK without `theme_void()`. There is ongoing github discussions with some of the developers at `ggplot2` to figure out why, but for now just make sure you use `theme_void` to add shapefiles to the same plot in R and have them work properly. Also take note of the `scale_fill_distiller` function. This is an easy way to throw in a `RColorbrewer` color pallette. In this case we used the Blues colorpallete since we're dealing with water but you can make it any `RColorbrewer` pallette you want. Also notice the `direction=1`. This reverses the order of the blues color pallette. I can't explain why it's not `-1`...Sidenote - you can also use `scale_fill_viridis()` which has some different default pallettes such as - “magma”, “plasma”, and “inferno”. 



#### Mapview
Saving the easiest for last...`mapview` is an awesome package that I didn't find until recently. As always though, a house with a foundation is more sound than one without. It's good to know about the other packages above!



```{r}
library(mapview)
mapview(merged.coast['maxpred'], col.regions = mycolours)
```




#### Saving a new shapefile
So we combined a csv with a shapefile to make that plot. We don't need to duplicate that hard work again because we can actually save our merged shapefile/csv as an ENVI shapefile.

`writeOGR(obj, dsn, layer, driver, dataset_options = NULL,
 layer_options=NULL, verbose = FALSE, check_exists=NULL,
 overwrite_layer=FALSE, delete_dsn=FALSE, morphToESRI=NULL,
 encoding=NULL, shp_edge_case_fix=FALSE)`
 

```{r, eval=FALSE}
writeOGR(obj = merged.coast, dsn = "/Users/james/Downloads/coast_files/merged.coast", layer = "coast-rgdal", driver = "ESRI Shapefile")
```



# Week 5 Assignment - Replicate these Plots

Using the **ne_10m_parks_and_protected_lands** shapefile dataset, map the names of the protected lands in the **Pacific Northwest** using the methods listed below. You will need to download all of the contents of the **ne_10m_parks_and_protected_lands** shapefile folder. *Note that you have to fuse a csv and to this shapefile*

1) spplot
2) tmap
3) ggplot2
4) mapview
 

Create the 4 plots above and make them look as close as possible to the ones below. 



```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=10}
library(maptools)
m <- readOGR("/Users/james/Documents/Github/geog473-673/datasets//ne_10m_parks_and_protected_lands/ne_10m_parks_and_protected_lands_area.shp")

spplot(m,"name", par.settings = list(axis.line = list(col ="transparent")), main = "Protected Lands", cuts = 5, col ="transparent",
       xlim = c(-127, -110), ylim = c(40,50), fill="darkgreen", colorkey=FALSE)


library(tmap)
tm_shape(m,bbox=tmaptools::bb(matrix(c(-127,40,-110,50),2,2))) + 
  tm_polygons(col='name', title = "Protected Lands", palette = "Spectral") + tm_style("classic") + tm_scale_bar(position = c("right", "bottom")) +
  tm_layout(legend.title.size = 0.7,legend.text.size = 0.5, legend.position = c("left","center")) +
  tm_layout(title="Protected Areas in Pacific NW")


states <- readOGR("/Users/james/Documents/Github/geog473-673/datasets/ne_10m_admin_1_states_provinces/ne_10m_admin_1_states_provinces.shp")
class(states)
states = st_as_sf(states)
class(states)








ggplot(data = states) + geom_sf() + theme_void() +
  geom_sf(data = st_as_sf(m), aes(fill=name), fill="darkgreen", colour = "aquamarine3", alpha=0.8) +
  coord_sf(xlim = c(-127, -110), ylim = c(40,50), expand = FALSE) + 
  theme(legend.position = "none") + ggtitle("Protected Lands in Pacific NW")







library(mapview)
mapview(m['name'], col.regions = mycolours)


```