---
title: "Week5 - Reprojecting and plotting ggplot2"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reprojecting Raster data and plotting it with ggplot2

### Quick tutorial - ggplot2 themes
Ever notice the gray background on ggplot2? Well that's referred to as the default theme. There are other themes that can be called every time you createa a ggplot. Let's take a look:

The black & white theme
```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors",     x="Year", caption="National Park Database") +
    theme_bw()
  
```

The dark theme

```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors",     x="Year", caption="National Park Database") +
    theme_dark()
  
```

The light theme

```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors",     x="Year", caption="National Park Database") +
    theme_light()
  
```

The classic theme

```{r}
library(ggplot2)

acadia = read.csv("/Users/james/Documents/Github/geog473-673/datasets/acadia.csv")

ggplot(acadia, aes(x=year, y=visitors)) +
    geom_line(col="yellow", size=5) + 
    geom_point(col="steelblue", size=3) + 
    geom_smooth(method="lm", col="firebrick") +
    labs(title="Acadia National Park Attendance", subtitle="Total Visitors per year", y="Visitors",     x="Year", caption="National Park Database") +
    theme_classic()
  
```


# Reprojecting Raster data and plotting it with ggplot2
