---
title: 'Week2: R Syntax, Classes, & Datatypes'
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome Back!

Welcome to the Advanced phase of R. In this class, we'll be covering new topics at an accelerated pace. But first, let's shake off the rust
and get back into the swing of things. As always, if you need revisit older tutorials, they can be found here - https://github.com/jsimkins2/geog473-673/tree/master/R

## Some Useful Definitions

| Function                 |             Description        |
| -----------| -------------------------------------------- |
| ls | lists contents of R workspace/global environment | 
| rm | removes objects from R workspace | 
| save|  save selected objects |
| +,-,*,/,^ | arithmetic operators | 
| %*% |  matrix multiplication |
| t | matrix transpose | 
| solve | matrix inverse (and solving linear equations) | 
| c | combines (concatenates) objects, simplest way to make vectors | 
| seq | creates vectors that are regular sequences | 
| rep | replicates vectors | 
| length| returns length of a vector | 
| sum | returns the sum |
| mean | returns the mean | 
| median | returns the median | 
| sd | returns the standard deviation (n − 1 in denominator) | 
| min | returns minimum | 
| max | returns maximum |
| sort | sort a vector (rearranges the vector in order) |
| order | returns indices of vectors that will order them |
| rank | returns rank of each element in vector | 
| ==, <, > | comparison operators |
| <=, >=, != | |
| |, & | OR, AND |
| is.na | tests for missing value NA | 
|which |does logical comparison and indicates which elements are TRUE that is, gives the TRUE indices of a logical object|
| any |does logical comparison returns 1 (TRUE) if any of the comparisons are TRUE, i.e. is at least one of the values true? |
|exp |returns e to that power|
|log |returns natural logarithm (to the base e)|
|log10| returns logarithm (to the base 10)|
|sqrt| returns square root|
|table |does frequencies and cross-tabs|
|help |help page on specified function|
|cbind |combine by columns|
|rbind |combine by rows|
|matrix| create a matrix|
|vector |create a vector|
|nrow| number of rows in an array or data frame|
|ncol |number of columns in an array or data frame|
|dim| dimensions of an array or data frame|
|array| create an array|

| Function                 |             Description        |
| -----------| -------------------------------------------- |
|is.vector |answers the question, is this a vector TRUE or FALSE|
|as.vector |attempts to coerce object into a vector|
|read.table |reads data from a text file|
|read.csv |reads data from a text file with comma separated data|
|write.table| writes a data frame to a text file|
|is.data.frame| tests object to see if it is data frame|
|as.data.frame |coerces object into data frame|
|is.factor |tests object to see if it is a factor|
|as.factor |coerces object into a factor|
|head, tail |list the first, last six rows|
|names |returns names of elements of object|
|colnames| returns or sets column names of object|
|rownames |returns or sets row names of object|
|subset| select part of a vector, matrix, or data frame|
|merge |merge two data frames|
|lm |multiple linear regression|
|glm |generalized linear regression|
|anova |analysis of variance|
|chisq.test |Pearson’s Chi-squared test for count data|
|summary| shows results of various model fitting functions|
|predict |predicted results from model|
|hist| histogram|
|boxplot| box plot|
|plot |scatterplot|
|lines |connects points sequentially with lines (added to a plot)|
|segments| add lines to a plot (between pairs of points)|
|text |add text to a plot|
|legend |add a legend to a plot|
|abline |add a line to a plot by specifying its slope and intercept|
|passing |an lm object will result in adding the predicted line to the plot|
|x11 |open another graphics window (PC)|
|pdf |open a pdf file for recording graphics|
|dev.off| close graphics device|
|par(mfrow) |arranges multiple plots on same page (by row)|
|sample |produces a random sample of the specified values|
|set.seed |sets seed for next random sample (repeat random sample)|
|rnorm |produces a random sample from a normal distribution|
|qnorm |quantiles (percentiles) of normal distribution|
|pnorm |CDF of normal distribution|
|dnorm |PDF of normal distribution|
|rbinom |produces a random sample from a binomial distribution|

## Let's do some basic recap

Variable Assignment & Operations
```{r}
x = 7
y = 9
z <- y * (x/2 + 7) + sqrt(y)

print(x)
y
z

X <- 1:20
Y <- (1 + 1/X) + 2
X
Y

plot(X,Y)
```

Now let's add in some extra arguments to the plot function to make it prettier
```{r}
plot(x=X,y=Y, type = "p", pch=c(24), col = 'yellow', bg='blue', cex = 1.6, xlab = "X", ylab = "Y")
title("X vs Y")
```

Having trouble remembering the arguments that the function `plot()` accepts? Same...so just run `?plot` for RStudio's built in helper.

Recap:

* type == type of plot - p stands for points
* pch == plotting ‘character’, i.e., symbol to use
* col == color
* bg == background color (only applicable for pch's between 21:25)
* cex == size of the points
* xlab == x label
* ylab == y label 

## In Class Exercise: 
Create a plot above using a sequence of X values where Y is the log of the X values. Use both lines and points, pch of 21, cex of 1.5, and colors of your choosing. 
```{r, include=FALSE}
X=1:10
Y=log(X)
plot(x=X,y=Y, type = "b", pch=c(24), col = 'yellow', bg='blue', cex = 1.6, xlab = "X", ylab = "Y")
title("X vs Y")
```

## Loading in Data
Let's say you're an ecologist who collected some tree data samples out in the field. Let's load this into R and explore the data. In order to proceed, you'll need to download **TreeData.csv** which is located here - (https://github.com/jsimkins2/geog473-673/tree/master/datasets/TreeData.csv)

```{r}
# Navigate to the location where your file is stored locally. Use read.csv() function to load the data
treedat <- read.csv("~/Downloads/TreeData.csv")
treedat
```

`treedat` is a data frame. As a reminder, a data frame is essentially a 2-dimensional array that contains a *combination* of vectors (columns of data) that are of the class; integer, numeric, character. This is *different* from a matrix which can only contain *1 type* of data.

```{r}
# let's set the rownames equal to the tree column
treedat <- read.csv("~/Downloads/TreeData.csv", row.names='tree')
# print treedat
treedat
# look at the data frame variable names 
names(treedat)
# print the dbh variable
treedat$dbh
# rename the dbh variable, but first let's be sure our index of the dbh variable is correct
colnames(treedat)[4]
# yep, dbh is the index number 4 
colnames(treedat)[4] = "tree.diameter"
treedat

# Now let's do some plotting 
par(mfrow=c(1,2)) ## create plot array of 1 row x 2 columns
par(cex.axis=0.8) ## shrinks the name size of the x axes. If we don't do this, not all the names in the boxplot show up
par(cex.main=0.7) ## shrinks the name size of the titles. If we don't do this, the titles don't fit in the window
# Use the his() function to plot a histogram 
hist(treedat$BarkThick, xlab= "Bark Thickness (cm)", main= "Histogram: Bark Thickness", col= "gray50")
boxplot(SapDepth ~ spp, data= treedat, ylab= "SapDepth", col= "darkslateblue", main= "Boxplot: Sapwood Depth by Species")

```

## Assignment:
Using the TreeData.csv above, complete the following:

1. Rename 'spp' variable to 'species'
2. barplot(summary(treedat$season))
