# Reprojecting & Writing Rasters

This week work on handling raster datasets that have undesirable projections. We'll reproject these datasets and then write them to a new datafile that we can use in the future. 

## Indexing Data

```{r}
matA=matrix(1:16,4,4)
matA
matA[2,3]
matA[c(1,3),c(2,4)]
matA[1:3,2:4]
matA[1:2,]
matA[,1:2]
matA[1,]
dim(matA)
```
##In Class Exercise: 

Starting with this code...
```{r}
matA=matrix(1:16,4,4)
```

Make this matrix....
```{r, echo=FALSE}
matA=matrix(1:16,4,4)
matA = matA * 2
matA[3,4] = 39
matA[2,] = 47
matA[1,1] = 1
matA
```


## Resampling and Reprojecting

This week we'll be working an example with **globalTemClim1961-1990.nc**. This is Global Temperature climatology from 1961 to 1990. We'll look at resampling this raster to a different size (resolution). Next, we'll reproject this dataset. Reprojection and resampling are a frequent task for spatial datasets because the Earth isn't flat (despite what your distant relative on Facebook might think). Earth's shape (oblate spheroid) presents challenging projection issues. 

```{r, warning=FALSE, message=FALSE}
# load in the packages
library(raster)
library(rasterVis)
library(maptools) # also loads sp package

# load in dataset directly via raster package, specify varname which is 'tem' for 'temperature' 
temClim = raster("../datasets/globalTemClim1961-1990.nc", varname = 'tem', band=1)
temClim

# Create a new, blank raster that has a totally different sizing
newRaster = raster(nrow = 180, ncol = 360)
newRaster

#resample the temClim raster to the resizedRaster
resTemClim = resample(x=temClim, y=newRaster, method='bilinear') # can be set to nearest neighbor using 'ngb' method
resTemClim

#define new projection as robinson via a proj4 string. Note that this can also be achieved
# using EPSG codes with the following - "+init=epsg:4326" for longlat
newproj <- CRS("+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" )
newproj

# reproject the raster to the new projection
projTemClim = projectRaster(resTemClim,crs=newproj)
projTemClim

data(wrld_simpl)
plt <- levelplot(resTemClim, margin=F, par.settings=BuRdTheme,
                 main="January Global Average Temp 1961-1990")
plt + layer(sp.lines(wrld_simpl, col='black', lwd=0.4))

# convert the wrld_simpl land polygons to the robinson projection
wrld_simpl = spTransform(wrld_simpl, CRS("+proj=robin +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" ))

plt <- levelplot(projTemClim, margin=F, par.settings=BuRdTheme,
                 main="January Global Average Temp 1961-1990")
plt + layer(sp.lines(wrld_simpl, col='black', lwd=0.4))
```

## PNGs
The `png()` function is a function that saves a plot to png. After we invoke the function and fill out the arguments, we need to execute the plot code between the `png()` function and `dev.off()`. `dev.off()` tells R that you're done adding things to the plot and that it can be done plotting.


```{r, eval=FALSE}
png(filename = "~/Downloads/myPNG.png", width = 10, height = 6, units = 'in',res=100)
plt <- levelplot(projTemClim, margin=F, par.settings=BuRdTheme,
                 main="January Global Average Temp 1961-1990")
plt + layer(sp.lines(wrld_simpl, col='black', lwd=0.4))
dev.off()
```

## Writing Rasters

```{r, eval=FALSE}
writeRaster(x=projTemClim, filename="~/Downloads/projectedTemClim1961-1990.tif", format='GTiff',
            varname="Temperature", longname="Global Average Temperature January 1960-1990",
            xname="lon", yname="lat")

```

You can save these rasters in a variety of formats. If you're interested in looking them up, run `help(writeRaster)` and read about the `format` argument. 

## Assignment

1. Load in globalTemClim1961-1990.nc 
2. Extract data for January and July
3. Find difference between two months globally
5. Enhance resolution 2x using nearest neighbor method 
  - (*hint: run help(resample) if you get stuck*)
6. Plot in mollwide projection
  - (*"+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"*)
7. Write raster to NetCDF
8. Upload PNG **and** netCDF file to Canvas under week 5 assignment

``` {r end of reprojecting, echo = FALSE, message = FALSE, error = FALSE, warning = FALSE, fig.height=10, fig.width=16}
library(raster)
library(rasterVis)
library(maptools) # also loads sp package

# load in dataset directly via raster package, specify varname which is 'tem' for 'temperature' 
jan = raster("../datasets/globalTemClim1961-1990.nc", varname = 'tem', band=1)
july = raster("../datasets/globalTemClim1961-1990.nc", varname = 'tem', band=7)

temClim = july - jan
# Create a new, blank raster that has a totally different sizing
newRaster = raster(nrow = 2*nrow(temClim), ncol = 2*ncol(temClim))


#resample the temClim raster to the resizedRaster
resTemClim = resample(x=temClim, y=newRaster, method='ngb')


#define new projection as robinson via a proj4 string. Note that this can also be achieved
# using EPSG codes with the following - "+init=epsg:4326" for longlat
newproj <- CRS("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs" )


# reproject the raster to the new projection
projTemClim = projectRaster(resTemClim,crs=newproj)


data(wrld_simpl)

# convert the wrld_simpl land polygons to the robinson projection
wrld_simpl = spTransform(wrld_simpl, CRS("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"))

plt <- levelplot(projTemClim, margin=F, par.settings=BuRdTheme,
                 main="July- January: Global Average Temp 1961-1990")
plt + layer(sp.lines(wrld_simpl, col='black', lwd=0.4)) + layer(panel.text(0,0,"Prof Simkins", alpha=1, col = "black"))
```
