# Oceanography & Global Tree Cover

This week we'll use a different dataset from the GOES-16 satellite to explore new coding practices. We'll take what we have learned from the example and apply it to a global tree cover dataset. First, let's cover an important skill - understanding the `paste0()` function. 

## `Paste0()`

Last week we covered the for loop which is an essential programming skill. We also touched on the `paste0()` function, but this is incredibly useful and should be expanded upon. `paste0()` is the sister function of `paste()`, who's cousin is `sprintf()` . It depends on preference, but most coders I know gravitate towards `paste0()`. Let's check out all 3. 

``` {r, warning=FALSE}
# Paste Example 1 - default sep (aka separation) is space
paste("file", "number", "32")

# Paste Example 2 - set sep to "_"
paste("file", "number", "32", sep = "_")

# Paste0 Example 1 - 0 for 0 separating characters
paste0("file", "number", "32")

# sprintf example 1
sprintf("%s %s %s", "file", "number", "32")

# sprintf example 2
person <-"Grover"
action <-"flying"
message(sprintf("On %s I realized %s was...\n%s by the street", Sys.Date(), person, action))


# Notice that paste() is limiting because the separating character is not always present between
# each string you're concatenating
# Let's use paste0 here
fileList <- c('filename1', 'filename2', 'filename3', 'filename4')
dateFolder <- c('0813', '0814', '0815', '0816')
homeDir <- "~/Documents/"
pathList <- list()

for (i in 1:length(fileList)){
  print(i)
  tempString <- paste0(homeDir, dateFolder[i], '/', fileList[i])
  pathList[i] <- tempString
}

pathList

pathList[[1]]


# Now let's rewrite using sprintf
fileList <- c('filename1', 'filename2', 'filename3', 'filename4')
dateFolder <- c('0813', '0814', '0815', '0816')
homeDir <- "~/Documents/"
pathList <- list()

pathList <- sprintf("%s%s/%s", homeDir, dateFolder, fileList)

pathList

pathList[[1]]

#sprintf is a very useful and quick function for combining various strings into longer strings
#but paste0 allows you a little more freedom and is slightly more intuitive

```

Debug and fix the following code:

``` {r, eval=FALSE}
# Let's use paste0 here
fileList <- c('sstData20190912', 'sstData20190913', 'sstData20190914', 'sstData20190915')
dateFolder <- seq(0912, 0915, 1)
homeDir <- "~/Documents/"
pathList <- list()

for (i in fileList){
  print(i)
  tempString <- paste0(homeDir, dateFolder[i], '/', i)
  pathList[i] <- tempString
}

pathList

pathList[[1]]
```

```{r, echo=FALSE, eval=FALSE}
#Answer
fileList <- c('sstData20190912', 'sstData20190913', 'sstData20190914', 'sstData20190915')
dateFolder <- seq(0912, 0915, 1)
homeDir <- "~/Documents/"
pathList <- list()

for (i in seq_along(fileList)){
  print(i)
  tempString <- paste0(homeDir, '0', dateFolder[i], '/', fileList[i])
  pathList[i] <- tempString
}

pathList

pathList[[1]]
```
The final `pathList[[1]]` should look like this...

`"~/Documents/0912/sstData20190912"`


## Rasters

Rasters offer us a 2 dimensional dataset (matrix) of data with geospatial coordinates. Rasters can come in a variety of filetypes including netCDF, GeoTIFF, CSV (if they come in CSV, they'll be massive - this is an inefficient datatype), etc. We'll be looking at the GOES-16 Rolling 1 Day average datafile - **GOES_R_ROLLING_1DAY_20190814.nc**. The GOES-16 Satellite records sea surface temperatures every hour. A rolling 1 day is an average of the previous 24 hours of record of sea surface temperature (and this dataset is created by CEMA here at UD). 

``` {r, message=FALSE, warnings=FALSE}
library(maptools) # also loads sp package
library(sp)
library(ncdf4)
library(raster)
library(rasterVis)
library(RColorBrewer)
ncFile <- nc_open("/Users/james/Documents/Github/geog473-673/datasets/GOES_R_ROLLING_1DAY_20190814.nc")
sstGoes <- ncvar_get(ncFile, varid = "sst")
sstRast <- raster(sstGoes)

# plot using quick 'image' function from raster
image(sstRast)

# netcdf stores data backwards, so to convert it to raster we need to transpose it
sstRast <- t(sstRast)

# plot using quick 'image' function from raster
image(sstRast)

# now we see that the data is just upside down, so we use the flip() function
sstRast <- flip(sstRast, 2)

# plot using quick 'image' function from raster
image(sstRast)

sstRast

# notice that the extent and crs information are wrong/not included so let's fix this
lat = ncvar_get(ncFile, "latitude")
lon = ncvar_get(ncFile, "longitude")

# define new extents for the raster
extent(sstRast) = c(min(lon), max(lon), min(lat), max(lat))
# define the proj4 projection string
crs(sstRast) = "+proj=longlat +datum=WGS84 +no_defs "

sstRast
image(sstRast)

# load raster Visualization package
library(rasterVis)

# levelplot the sstRast
levelplot(sstRast)

# USA shapefiles via the getData function
usa <- getData('GADM', country = 'USA', level = 1)

# Throw together the usa spatial polygons data frame
plt <- levelplot(sstRast, margin=F, par.settings=BuRdTheme,
       main="GOES-R Rolling SST 08/14")
plt + layer(sp.polygons(usa, col='black',fill='grey', lwd=0.4))

# Now let's do the same with geotiff
tifFile = "/Users/james/Documents/Github/geog473-673/datasets/goesSST.tif"
sstRast = raster(tifFile)

# Quick image using raster's image plot generator function
image(sstRast)

# Now let's load a base dataset of world maps that R already has stored
data(wrld_simpl)
plt <- levelplot(sstRast, margin=F, par.settings=BuRdTheme,
       main="GOES-R Rolling SST 08/14")
plt + layer(sp.lines(wrld_simpl, col='black', lwd=0.4))

# Let's do the same plot but with a custom shapefile
# note that even though we only point the function to the '.shp' file, the '.shx' and '.dbf' need to be in the same folder for this to work. 
world.shp <- rgdal::readOGR("/Users/james/Documents/Github/geog473-673/datasets/world_shpfiles/world.shp")

# add custom color theme using brewer.pal from the package RColorBrewer and setting the rasterTheme
mapTheme <- rasterTheme(region=brewer.pal(8,"Reds"))

plt <- levelplot(sstRast, margin=F, par.settings=mapTheme,
       main="GOES-R Rolling SST 08/14")
plt + layer(sp.lines(world.shp, col='gray', lwd=0.4))


```

## Assignment

1. Download treecov.nc from the datasets folder

2. Open tree cover % as a variable, remove bad values.

3. Plot tree cover variable using a green theme. Be sure to add coastlines via your choice of underlying dataset. Ensure correct latitude/longitudes are displayed. Add title.

4. Submit resulting image to Canvas assignment 4