# Scientific R Scripts

So far, the examples of R scripts we've shown have been general purpose scripts. For this week, we're going to cover an example where we want to plot sea surface temperature from a given netCDF file. A netCDF file (network Common Data Format) is an advanced filetype that is incredibly efficient at storing data and heavily used in physical sciences. Within each file, metadata (time, latitude info, longitude info, projection, etc.), and variables (sea surface temperature, latitude points, longitude points, chlorophyll, etc.) can be found. The netCDF file we'll be looking at in this example is **OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc**. This is a data file from the GOES-16 (Geostationary Operational Environmental Satellite) satellite that shows a snapshot of the Northwest Atlantic Ocean and Eastern Seaboard. So we have the data, but what does it look like?

A scientific R Script might look like this:

```{r, message=FALSE, warning=FALSE, fig.align="center", out.width="60%"}
# James Simkins
# Load libraries
library(ncdf4)
library(fields)

# open netcdf file
ncFile <- ncdf4::nc_open("~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")

# extract SST - sea surface temperature 
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
lat <- ncdf4::ncvar_get(nc=ncFile, varid="latitude")
lon <- ncdf4::ncvar_get(nc=ncFile, varid="longitude")
# convert sst from Kelvin to Celsius
sstC <- sstK - 273.15
# remove values below 0C
sstC[sstC < 0] = NA
# Plot the matrix
fields::image.plot(x=lon, y=lat, z=sstC, legend.lab="Celsius")
title("GOES-R SST Product 7/25/2019 19:00 UTC")
```

## Script Breakdown

```{r, message=FALSE, warning=FALSE}
# Load libraries
library(ncdf4)
library(fields)
```

In R, we need to call on packages/libraries that we want to load in. As a reminder, packages are the toolbox we want to bring into R and the functions inside each package are the tools within that specific toolbox. The `library()` base function loads packages. After we load a package, R will know what we mean when we call on a function from that package. 

```{r, message=FALSE, warning=FALSE}
# Load libraries
library(ncdf4)
library(fields)
#######################################

ncFile <- ncdf4::nc_open(filename="~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
lat <- ncdf4::ncvar_get(nc=ncFile, varid="latitude")
lon <- ncdf4::ncvar_get(nc=ncFile, varid="longitude")
```

Our object name is `ncFile` and it holds an opened NetCDF file. We open this file via the `nc_open()` function that's within the `ncdf4` library. Note the `ncdf4::nc_open()` syntax. Using this syntax, we explicitly tell R we wish to use the `nc_open()` function within (`::`) the `ncdf4` package. This is not necessary after loading the `ncdf4` package above but is still important ot know.

Notice we can use either `<-` or `=` for objects (like `ncFile` or `sstK`) but I MUST use `=` within the *function* 'walls' (the parantheses). 

 We open the netcdf file and then extract what we want out of it using `ncvar_get()`, which is short for "netcdf variable get". Confused about how to use `ncvar_get()`? Try running `help(ncvar_get)` in your console. 


```{r, message=FALSE, warning=FALSE}
# Load libraries
library(ncdf4)
library(fields)
ncFile <- ncdf4::nc_open(filename="~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
#######################################
# convert sst from Kelvin to Celsius
sstC <- sstK - 273.15
```

The SST variable from the netCDF file was in Kelvin and we want to convert it to Celsius. Right now, sstK is a matrix. How do I know this? Look at your environment, or simply type into your console:

```{r, message=FALSE, warning=FALSE}
# Load libraries
library(ncdf4)
library(fields)
ncFile <- ncdf4::nc_open(filename="~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
#######################################

class(sstK)
```

Class is a useful function that is loaded with the base library everytime you fire up R. It tells us what type of object we have. Now that we know this is a matrix, we can subtract 0 Celsius, or 273.15 Kelvin. When we have a matrix in R and perform any math on it, it does that math on each and every matrix value.

#Quick and Dirty Quality Control

There are bad values that crept into the dataset and we need to convert all of them to `NaN` (aka Not A Number...also known as `NA` (Not Available) in R). How do we know there are bad values in this dataset?

```{r, message=FALSE, warning=FALSE}
# Load libraries
library(ncdf4)
library(fields)
ncFile <- ncdf4::nc_open(filename="~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
#######################################

summary(as.vector(sstK))
```

`summary()` is another great base function. In order to use it on a matrix, we need to convert it to a vector - summary can't do 2 dimensional objects like matrices, it needs a one dimensional vector of numbers. Notice the `Min` in the summary output. -999? No way is that a valid Kelvin value, especially since we subtract another 273.15 to this number to make the the Celsius matrix. So clearly we have some bad data that we need to convert to `NA`'s. We do this by...

```{r, message=FALSE, warning=FALSE}
library(ncdf4)
library(fields)
ncFile <- ncdf4::nc_open("~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
lat <- ncdf4::ncvar_get(nc=ncFile, varid="latitude")
lon <- ncdf4::ncvar_get(nc=ncFile, varid="longitude")
# convert sst from Kelvin to Celsius
sstC <- sstK - 273.15
#######################################

# remove values below 0C
sstC[sstC < 0] = NA
```

This line reads as: `sstC` where `sstC` is less than `0` equals `NA`. The brackets here can be thought of as the 'condition', that is what you're looking to change. This is called a vector operation, which we will get more into later but these are important becuase it's far faster to do this than a for loop. 

```{r, message=FALSE, warning=FALSE}
library(ncdf4)
library(fields)
ncFile <- ncdf4::nc_open("~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
lat <- ncdf4::ncvar_get(nc=ncFile, varid="latitude")
lon <- ncdf4::ncvar_get(nc=ncFile, varid="longitude")
# convert sst from Kelvin to Celsius
sstC <- sstK - 273.15
# remove values below 0C
#######################################

sstC[sstC < 0] = NA
summary(as.vector(sstC))
```

Now we see that our summary looks good (even though we're looking at a matrix in celsius now). So all that's left to do is plot this up...

```{r, message=FALSE, warning=FALSE, fig.align="center", out.width="60%"}
# James Simkins
# Load libraries
library(ncdf4)
library(fields)
ncFile <- ncdf4::nc_open("~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(nc=ncFile, varid="SST")
lat <- ncdf4::ncvar_get(nc=ncFile, varid="latitude")
lon <- ncdf4::ncvar_get(nc=ncFile, varid="longitude")

# convert sst from Kelvin to Celsius
sstC <- sstK - 273.15
# remove values below 0C
sstC[sstC < 0] = NA

#######################################
# Plot the matrix
fields::image.plot(x=lon, y=lat, z=sstC, legend.lab="Celsius")
title("GOES-R SST Product 7/25/2019 19:00 UTC")
```


Now we plot this up using the `image.plot()` function from the `fields` library. `image.plot()` takes the arguments `x`, `y`, and `z`. This is a 2 dimensional dataset of Sea Surface Temperature where the `x` is `longitude`, the `y` is `latitude`, and what we plot within the box (`z`) is `sstC`. Notice that for this kind of plot, the 'title()' function is a separate function rather than an argument that goes inside of `image.plot()`. Remember, if you're ever confused about something like this just tell R you need 'help()'. 


## In Class Exercise

1) Go to https://github.com/jsimkins2/geog473-673/tree/master/datasets
2) Download 'OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc'
3) Make simple plot of the `DQF` (Data Quality Flag) variable

-Hint: You may have to use `install.packages()` to install pacakages you may not have yet on your machine.

Aim for something like this:
```{r, echo=FALSE, fig.align="center", out.width="60%"}
library(ncdf4)
library(fields)
ncFile <- ncdf4::nc_open("~/Documents/Github/geog473-673/datasets/OR_ABI-L2-SSTF-M3_G16_s20192081300453_e20192081400161_c20192081406297.nc")
sstK <- ncdf4::ncvar_get(ncFile, "DQF")
lat <- ncdf4::ncvar_get(ncFile, "latitude")
lon <- ncdf4::ncvar_get(ncFile, "longitude")

#png("~/Documents/Github/geog473-673/R-Section1/Week2_RSyntaxClassesDatatypes_files/figure-markdown_github/Week2_dqf_plot.png", width = 10, height = 6, units = "in",res = 200)
fields::image.plot(x=lon, y=lat, z=sstK, legend.lab = "Level")
title("GOES-R DQF 7/25/2019 19:00 UTC")
#dev.off()
```

### Further Explanation of DataTypes

If you look into your environment from the in class exercise, you'll notice under the 'Data' tab you have a large matrix of the DQF values. R stores these matrices without the column/row identifier (aka lat & lon). We provide the image.plot() function with the lon & lat arrays becuase it doesn't know what the x & y coordinates are of the matrix. Notice that the environment tells you the dimensions - the matrix is 2778 x 1989. R is indexed from 1 to the length of the dimension. Here is what I mean:

- ```dim(sstC)``` is ```2778 1989```
- ```dim(lat)``` is ```1989```
- ```lat[0]``` is ```numeric(0)```
- ```lat[1]``` is ```16.00283```
- ```lat[1989]``` is ```51.98563```
- ```lat[1990]``` is ```NA```

The above tests are referred to as indexing. The 1st point of the lat array is 16.00283. In R, we index using brackets []. If you want to find more values other than just a single point, the procedure is referred to as indexing/slicing/subsetting.

```lat[1:10]```
is
``` 16.00283 16.02093 16.03903 16.05713 16.07523 16.09333 16.11143 16.12953 16.14763 16.16573```

The lat object we've been exploring here is an 'array'. 

An array is a vector with one or more dimensions. So, an array with one dimension is (almost) the same as a vector. An array with two dimensions is (almost) the same as a matrix. An array with three or more dimensions is an n-dimensional array.

A vector is what is called an array in all other programming languages except R — a collection of cells with a fixed size where all cells hold the same type (integers or characters or reals or whatever).
