---
  title: 'Week3: Time Series fun with DEOS'
output: rmarkdown::github_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General Skills: The For Loop

For loops & conditional statements are a key skill in programming. They allow you to process through large datasets or multiple datasets thus minimizing the amount of manual work you need to do. The basic for loop looks like this...

```{r}
# Generate sequence of numbers from 1 to 10 using the seq() function (seq for sequence)
numbersList = seq(from=1,to=10,by=1)
# Multiply each number in the numbersList by 8 and print the result
for (i in numbersList){
  temNumber = i * 8
  print(temNumber)
}
```

Notice the general structure of R for loops. 'for' signals to R you're beginning a for loop, which requires the general structure to look like:

for (something in something inside these parentheses){
  do something within these curly brackets
}

Yes, you must have these parentheses and curly brackets present and surrounding the appropriate code. If you forget a parentheses or curly bracket you'll have errors pop up...this happens to me all the time still. 

While these must be present, R doesn't care where they are in your code (****which is very unique amongst programming languages). For example, notice how this ugly code is different but still runs...

```{r}
# Generate sequence of numbers from 1 to 10 using the seq() function (seq for sequence)
numbersList = seq(from=1,to=10,by=1)
# Multiply each number in the numbersList by 8 and print the result
for (        i in numbersList       
      ){
  temNumber = i * 8
  print(temNumber)
        }
```


The general structure is still: for(condition){do something}. If statements are set up the same way

```{r}
# Generate sequence of numbers from 1 to 10 using the seq() function (seq for sequence)
numbersList = seq(from=1,to=10,by=1)
# Multiply each number in the numbersList by 8 and print the result
for (i in numbersList){
  if (i==4){
    temNumber = i * 8
    print(temNumber)
  }
}
```

This is referred to as a 'nested loop', because there is a conditional statement within another one. Key takeaway here: in programming languages, '=' is an assignment (i.e. x = 4), whereas '==' is an equality test (i == 4). To put this loop in layman's terms: for i in numbersList, if i is equal to 4, multiply i by 8 and then print temNumber. 

We can also have nested for loops. 

```{r}
# Generate sequence of numbers from 1 to 3 this time using the seq() function (seq for sequence)
numbersList = seq(from=1,to=3,by=1)
lettersList = list("A", "B", "C")

for (num in numbersList){
  for (let in lettersList){
    print(c(num,let))
  }
}
```

You can name the object within the list whatever you want (i, j, num, let, etc.). Also, c() is the concatenate functin that combines values into a vector or list. The order doesn't matter in this for loop...

```{r}
# Generate sequence of numbers from 1 to 3 this time using the seq() function (seq for sequence)
numbersList = seq(from=1,to=3,by=1)
lettersList = list("A", "B", "C")

for (let in lettersList){
  for (num in numbersList){
    print(c(num,let))
  }
}
```

But it does in this one...

```{r}
# Generate sequence of numbers from 1 to 10 using the seq() function (seq for sequence)
numbersList = seq(from=1,to=10,by=1)
# Multiply each number in the numbersList by 8 and print the result
if (i==4){
  for (i in numbersList){
    temNumber = i * 8
    print(temNumber)
  }
}
```

Here's one more example for multi conditional statement with an else...

```{r}
# Generate sequence of numbers from 1 to 3 this time using the seq() function (seq for sequence)
numbersList = seq(from=1,to=3,by=1)
lettersList = list("A", "B", "C")

for (num in numbersList){
  for (let in lettersList){
    if (num == 3 && let == "B"){
      print(c(num,let))
    } else{
      print("Not what we want")
    }
  }
}
```

&& means "and" ... || means "or"...these are useful in multi conditional statements. The 'else' statement is an appendage of the 'if' statement. It basically means if num == 3 and let == B is false, print "not what we want". Notice that the 'else' statement is outside of the 'if' statement but immediately after it. 


## DEOS Data

In Class Exercise: debug a for loop 

In Class Exercise: write DEOS exploration code together, show basic plot

Wednesday Class: Dr. Leathers Prompt. Using DEOS data, go quality control it, do a basic plot 

Why do we do quizzes? Because harnessing these names will make google searching easier

